Vagueness and imprecise imitation in signaling games

### 2016-02-22 paper/version_05/paper.tex
introduction extended

@@

But there are other intriguing aspects about vagueness. The puzzle
that we are concerned with here is how vagueness could arise and be
maintained in the first place. This is a serious issue for
functionalist accounts that maintain that concepts and linguistic
meanings evolved driven towards efficiency. Since the existence of
unclear borderline cases seems to entail inefficiency of
categorization or communication, the challenge, succinctly put forward
by \citet{Lipman2009:Why-is-Language}, is to explain how vagueness can
persist under evolutionary pressure to be optimally discriminative. We
address \emph{Lipman's problem} as a technical challenge, namely to
find a conceptually sound and mathematically coherent formal model of
the evolution of categorization and language use that yields
by-and-large efficient categorization \emph{and} vagueness.

A number of authors have recently tried to explain why vagueness evolved as something that is
itself useful
\citep[e.g.][]{Jaegherde-Jaegher2003:A-Game-Theoreti,Deemter2009:Utility-and-Lan,BlumeBoard2013:Intentional-Vag}.
Others have argued that vagueness is a natural byproduct of limitations in information
processing \citep[e.g.][]{FrankeJager2010:Vagueness-Signa} or the result of particular
low-level learning strategies \citep[e.g.][]{OConnor2013:The-Evolution-o}. This paper
contributes to the latter line of thought. Concretely, we introduce a dynamic that describes
changes in aggregate population behavior as the result of imitation with probabilistic
confusability of similar stimuli. We show that stochastic noise can not only lead to vague, yet
relatively efficient signal use, but can also unify evolutionary outcomes and help avoid
suboptimal categorization: regularities in confusability of stimuli also regularize signaling
behavior and evolving categories. The system gives rise to generalization behavior without
agents actually generalizing.

The next section introduces the background against which the work presented here can be
appreciated. Section~\ref{sec:repl-diff-dynam} introduces a generalization of the replicator
dynamic which is derived from the idea that agents imitate other agents' behavior while
possibly confusing similar states. This imprecise imitation dynamic is explored in
Section~\ref{sec:exploring-rdd}. Section~\ref{sec:discussion} reflects and compares our
approach to others. Appendix~\ref{sec:formal-stuff} provides formal detail.

>>>>

\definecolor{Red}{RGB}{178,34,34}
\newcommand{\mf}[1]{\textcolor{Red}{[mf: #1]}}

@@

But there are other intriguing aspects about vagueness. One perplexing issue is how vagueness
could arise and be maintained in the first place. This is an apparent puzzle for functionalist
accounts that maintain that concepts and linguistic meanings evolved driven towards
efficiency. Since the existence of unclear borderline cases seems to entail inefficiency of
categorization or communication, the challenge, succinctly put forward by
\citet{Lipman2009:Why-is-Language}, is to explain two things at the same time: (i) how
vagueness can persist under evolutionary pressure to be optimally discriminative, (ii) without
jeopardizing the possibility of evolving, learning and communicating with a meaningful
language. A number of authors have consequently tried to explain why vagueness evolved as
something that is itself useful
\citep[e.g.][]{Jaegherde-Jaegher2003:A-Game-Theoreti,Deemter2009:Utility-and-Lan,BlumeBoard2013:Intentional-Vag}.
Others have argued that vagueness is a natural byproduct of limitations in information
processing \citep[e.g.][]{FrankeJager2010:Vagueness-Signa} or of generalization in low-level
learning strategies \citep[e.g.][]{OConnor2013:The-Evolution-o}. This paper contributes to the
latter line of thought. We believe that vagueness in language and thought may have many
reasons, not just one. We focus here on one possibility that has not received due attention,
namely imprecise perception of similar stimuli and imprecise realization of similar
responses.

Our conceptual starting point is the widely shared conviction that language is shaped by at
least two forces, which may, on occasion, pull in opposite direction. On the one hand, there is
functional pressure towards efficient communication. On the other hand, there is systematic
error, noise or imprecision in the transmission of linguistic behavior, knowledge or traits. As
an example for the latter, consider a child learning syntactic rules from a parent
generation. The child must infer these unobservable rules from observable speech. Inductive
biases of the learner may influence which syntactic rules are likely to be inferred from
(finite) parental input. Over the course of many generations, the effects of such transmission
biases can, in a manner of speaking, ``accumulate'' and lead to surprising results, such as the
evolution of compositional form-meaning mappings or the use of regular recursive syntactic
structure. This can happen when the effects of tranmission biases are isolated, as in iterated
learning models
\citep{KirbyHurford2002:The-Emergence-o,SmithKirby2003:Iterated-Learni,GriffithsKalish2007:Language-Evolut,KirbyGriffith2014:Iterated-Learni},
or when they interact with functional pressure towards efficient communication, e.g., as
formalized in the replicator mutator dynamic
\citep[e.g.][]{NowakPlotkin2000:The-Evolution-o,NowakKomarova2001:Evolution-of-Un}.

The emphasis of previous models that studied the effects of transmission infidelity on the
evolution of language has been on inductive biases and the systematicity, compression and
regularization that they can introduce. Here, we would like to show that transmission noise of
a different kind can lead to regularization as well and also give rise to vague
meaning. Towards this end, we formalize the expected change in the behavior of a population of
agents that try to imitate other agents' signaling behavior. We assume, however, that both
realization of behavior, as well as observability of others' behavior, is molested by noise, in
particular the probabilistic confusability of similar stimuli. The resulting population-level
dynamic generalizes the replicator dynamic \citep{TaylorJonker1978:Evolutionary-St} in its
interpretation as a cultural evolutionary dynamic based on imitation
\citep[e.g.][]{Helbing1996:A-Stochastic-Be,Schlag1998:Why-Imitate-and}. The inclusion of
confusability of stimuli does not undermine the possibility of evolving communicative signaling
behavior. Instead, it leads to the evolution of vague meanings. It moreover unifies and
regularizes evolutionary outcomes, making it appear as if agents were applying inductive biases
or generalizing over partial observations, when in truth this is the sole effect of confusion
of similar stimuli.

% The replicator mutator dynamic is neat because it keeps the effects of fitness-based selection
% (replication) and transmission biases (mutation) nicely apart. This is possible because it
% adheres to a picture of (asexual) biological reproduction: the number of children a (single)
% parent has is higher, the higher the communicative success of the parent's language; all
% children of this parent will learn from this parent alone, but may make mistakes during
% language acquisition.

The next section introduces the background against which the work presented here can be
appreciated. Section~\ref{sec:repl-diff-dynam} introduces our generalization of the replicator
dynamic which is derived from the idea that agents imitate other agents' behavior while
possibly confusing similar states. This imprecise imitation dynamic is explored in
Section~\ref{sec:exploring-rdd}. Section~\ref{sec:discussion} reflects and compares our
approach to others. Appendix~\ref{sec:formal-stuff} provides formal detail.

\mf{intro misses a mentioning of signaling games}
